---
project:
  title: "Project 1 - Linear Regression on Plasma β-carotene and Personal Factors"
author: 
  - name: "Mattis Ranheim"
    affiliation: "Lund University"
    email: "ranheim.mattis@gmail.com"
format:
  pdf:
    toc: true
    number-sections: true
    fig-cap-location: bottom
    keep-tex: false
    link-citations: true
    mainfont: "Palatino" 
execute:
  echo: false 
  warning: false
  message: false

editor: visual
---

# Project 1 - Report: Linear Regression

## Use of AI Tools

AI tools were used to assist in writing and coding: - **ChatGPT (OpenAI)** was used to clarify statistical concepts, draft parts of code, and suggest text structure. - All code was reviewed, understood, and adapted by the author. - Output was carefully verified for correctness.

Spelling and grammar suggestions from **RStudio Visual Editor** were used.

------------------------------------------------------------------------

## Author Contributions

| Name | Roles |
|--------------------------|----------------------------------------------|
| Mattis Ranheim | Derivations, Analysis, Discussions, Programming, Visualisation, Writing (original draft), Writing (revision & editing), Project Management |

# Introduction

Numerous observational studies have suggested that low dietary intake or low plasma concentrations of β-carotene and other carotenoids may be linked to an increased risk of developing certain types of cancer. However, relatively few studies have examined which factors actually influence plasma concentrations of these micronutrients.

In this project, we analyze data from a cross-sectional study conducted by Nierenberg et al. (1989), where the goal was to investigate the relationship between **personal characteristics**, **dietary intake**, and **plasma concentrations of β-carotene**. The study population consisted of 315 patients who underwent elective surgical procedures to biopsy or remove benign (non-cancerous) lesions in organs such as the lung, colon, breast, skin, ovary, or uterus. For this analysis, we focus exclusively on **plasma β-carotene concentrations** as the outcome of interest.

The study highlights considerable individual variation in plasma β-carotene levels and suggests that much of this variability may be explained by lifestyle and dietary factors.

# Data Description

The dataset used in this project contains **315 observations** and **12 variables**, stored in the file `carotene.xlsx`. Each row corresponds to an individual patient from the study. The variables are described below:

| Variable | Description |
|--------------------------------------|----------------------------------|
| `age` | Age (years) |
| `sex` | Sex (1 = Male, 2 = Female) |
| `smokstat` | Smoking status (1 = Never, 2 = Former, 3 = Current) |
| `bmi` | Body mass index (BMI = weight/height², kg/m²) |
| `vituse` | Vitamin use (1 = Yes, fairly often, 2 = Yes, not often, 3 = No) |
| `calories` | Daily calorie intake (MJ) |
| `fat` | Fat consumed per day (g) |
| `fiber` | Fiber consumed per day (g) |
| `alcohol` | Alcoholic drinks per week |
| `cholesterol` | Daily cholesterol intake (mg) |
| `betadiet` | Dietary β-carotene intake per day (mg) |
| `betaplasma` | **Plasma β-carotene concentration (ng/ml)** — this is the **response variable** we aim to model |

Our objective is to model how `betaplasma` varies as a function of the other variables using a **linear regression model** of the form:

$$
Y_i = \mathbf{x}_i \boldsymbol{\beta} + \varepsilon_i
$$

where $Y_i$ is the plasma β-carotene concentration for individual $i$, $\mathbf{x}_i$ is the vector of explanatory variables, $\boldsymbol{\beta}$ is the vector of unknown regression coefficients, and $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$ are the error terms.

To satisfy the linear model assumptions (e.g., normality and homoscedasticity of residuals), we may need to apply **suitable transformations** to the response and/or predictor variables throughout the analysis.

# 1. Testing Model Assumptions: Linear vs Log-transformed

We fitted two models to examine the relationship between BMI and plasma β-carotene levels:

-   **Linear model**: `betaplasma ~ bmi`
-   **Log-transformed model**: `log(betaplasma) ~ bmi`

The aim is to assess whether a log-transformation of the outcome variable improves model fit and better satisfies the assumptions of linear regression — particularly **normality of residuals** and **constant variance (homoscedasticity)**.

Below, we compare the two models using **residual plots** and **Q-Q plots** for both. A good model should show no patterns in the residuals vs fitted plot, and the residuals should lie close to the theoretical line in the Q-Q plot.

These graphs show the residual and QQ-plots for the **Linear** and log-trans

```{r}
#Loading the data
library(readxl)
data <- read_excel("data/carotene.xlsx") 
```

```{r}
# Linear model (no transformation)
model_lin <- lm(betaplasma ~ bmi, data = data)

# Log-transformed model
model_log <- lm(log(betaplasma) ~ bmi, data = data)
```

```{r}

library(ggplot2)
library(patchwork)

# Residuals
data$res_lin <- resid(model_lin)
data$res_log <- resid(model_log)
data$fit_lin <- fitted(model_lin)
data$fit_log <- fitted(model_log)

# Linear model plots
p1 <- ggplot(data, aes(x = fit_lin, y = res_lin)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted (Linear)", x = "Fitted", y = "Residuals") +
  theme_minimal()

p2 <- ggplot(data, aes(sample = res_lin)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot (Linear)", x = "Theoretical", y = "Sample Quantiles") +
  theme_minimal()

# Log model plots
p3 <- ggplot(data, aes(x = fit_log, y = res_log)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue") +
  labs(title = "Residuals vs Fitted (Log)", x = "Fitted", y = "Residuals") +
  theme_minimal()

p4 <- ggplot(data, aes(sample = res_log)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  labs(title = "Q-Q Plot (Log)", x = "Theoretical", y = "Sample Quantiles") +
  theme_minimal()

(p1 | p2) / (p3 | p4)

```

The residual plots and Q-Q plots show that the **log-transformed model** produces more homoscedastic residuals and better alignment with the normal distribution in the Q-Q plot. In contrast, the residuals of the linear model display signs of heteroscedasticity and heavier tails.

This suggests that the log transformation stabilizes the variance and brings the residuals closer to normality. Therefore, the **log-transformed model is more suitable** for satisfying the assumptions of linear regression.

### Model Estimates

To interpret the relationship between BMI and plasma β-carotene concentration, we present the coefficient estimates from the log-linear model:

The table below shows the **β-estimates** and their associated **95% confidence intervals**. The intercept corresponds to the expected value of log(β-carotene) when BMI is zero (which is not realistic in practice, but needed for the mathematical formulation), while the slope for BMI describes the expected **multiplicative change** in β-carotene concentration for each one-unit increase in BMI.

```{r}
# Fit log model
model_log <- lm(log(betaplasma) ~ bmi, data = data)

# Create table of estimates + 95% CI
conf_table <- as.data.frame(confint(model_log))
colnames(conf_table) <- c("2.5 %", "97.5 %")
conf_table$Estimate <- coef(model_log)
conf_table <- conf_table[, c("Estimate", "2.5 %", "97.5 %")]

knitr::kable(conf_table, digits = 4, caption = "β estimates and 95% confidence intervals for log-linear model")
```

The estimate for **BMI** is **−0.0359**, with a 95% confidence interval from **−0.0494 to −0.0224**, indicating a statistically significant negative association. This suggests that for every additional unit increase in BMI, the **log of plasma β-carotene decreases**, implying an **approximate 3.5% reduction** in β-carotene levels per BMI unit.

This negative association supports the hypothesis that higher body fat may be linked to lower concentrations of this micronutrient.

## Linear Model with Log Transformation and Back-Transformation

To investigate how plasma β-carotene levels relate to BMI, we fit a linear regression model where the outcome was **log-transformed β-carotene concentration**. This transformation helps satisfy linear regression assumptions, especially linearity and homoscedasticity.

Below, we show two plots:

-   The **top plot** shows the relationship between BMI and the log-transformed β-carotene levels, with fitted line, 95% confidence interval, and 95% prediction interval.
-   The **bottom plot** displays the same model but transformed back to the original β-carotene scale (ng/ml). This gives a more intuitive interpretation of the effect in absolute terms.

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=7, fig.cap="Log-scale and back-transformed plots with 95% CI and prediction intervals."}


# Prediction grid
bmi_seq <- data.frame(bmi = seq(min(data$bmi), max(data$bmi), length.out = 100))
pred_ci <- predict(model_log, newdata = bmi_seq, interval = "confidence")
pred_pi <- predict(model_log, newdata = bmi_seq, interval = "prediction")

# Combine into data frame
plot_df <- cbind(bmi_seq, pred_ci,
                 pi_lwr = pred_pi[, "lwr"],
                 pi_upr = pred_pi[, "upr"])

# --- Plot 1: log(betaplasma) vs BMI ---
plot_log <- ggplot() +
  geom_point(data = data, aes(x = bmi, y = log(betaplasma)), alpha = 0.6) +
  geom_line(data = plot_df, aes(x = bmi, y = fit), color = "blue") +
  geom_ribbon(data = plot_df, aes(x = bmi, ymin = lwr, ymax = upr), fill = "blue", alpha = 0.3) +
  geom_ribbon(data = plot_df, aes(x = bmi, ymin = pi_lwr, ymax = pi_upr), fill = "gray", alpha = 0.15) +
  labs(title = "Log(β-carotene) vs BMI",
       x = "BMI", y = "log(β-carotene)") +
  theme_minimal()

# --- Back-transform predictions ---
plot_df$fit_exp <- exp(plot_df$fit)
plot_df$lwr_exp <- exp(plot_df$lwr)
plot_df$upr_exp <- exp(plot_df$upr)
plot_df$pi_lwr_exp <- exp(plot_df$pi_lwr)
plot_df$pi_upr_exp <- exp(plot_df$pi_upr)

# --- Plot 2: betaplasma vs BMI (back-transformed) ---
plot_back <- ggplot() +
  geom_point(data = data, aes(x = bmi, y = betaplasma), alpha = 0.6) +
  geom_line(data = plot_df, aes(x = bmi, y = fit_exp), color = "blue") +
  geom_ribbon(data = plot_df, aes(x = bmi, ymin = lwr_exp, ymax = upr_exp), fill = "blue", alpha = 0.3) +
  geom_ribbon(data = plot_df, aes(x = bmi, ymin = pi_lwr_exp, ymax = pi_upr_exp), fill = "gray", alpha = 0.15) +
  labs(title = "Plasma β-carotene vs BMI (Back-transformed)",
       x = "BMI", y = "β-carotene (ng/ml)") +
  theme_minimal()

# Display side-by-side
plot_log / plot_back
```

The log-scale plot shows a clearer linear trend and narrower intervals, while the back-transformed version reveals the actual magnitude of plasma β-carotene and illustrates the asymmetry introduced by the exponential function.

These results suggest that BMI is negatively associated with β-carotene levels, and that using a log transformation is appropriate to model this relationship under the assumptions of linear regression.

### Estimated Changes in Plasma β-carotene for Changes in BMI

We interpret the log-linear regression model by expressing expected **percentage changes** in plasma β-carotene (ng/ml) for three different BMI changes:

1.  **BMI increased by 1 unit**
2.  **BMI decreased by 1 unit**
3.  **BMI decreased by 10 units**

```{r}
# Extract BMI coefficient and confidence interval
beta1 <- coef(model_log)["bmi"]
beta1_ci <- confint(model_log)["bmi", ]

# Function to compute percent change
pc <- function(b) (exp(b) - 1) * 100

# Compute estimates
change_table <- data.frame(
  `BMI Change` = c("+1", "−1", "−10"),
  `Estimate (%)` = round(c(pc(beta1), pc(-beta1), pc(-10 * beta1)), 2),
  `95% CI Lower (%)` = round(c(pc(beta1_ci[1]), pc(-beta1_ci[2]), pc(-10 * beta1_ci[2])), 2),
  `95% CI Upper (%)` = round(c(pc(beta1_ci[2]), pc(-beta1_ci[1]), pc(-10 * beta1_ci[1])), 2)
)

knitr::kable(change_table,
             caption = "Estimated percentage change in plasma β-carotene for selected changes in BMI (with 95% CI)")
```

From the table, we can conclude the following:

-   A **1-unit increase in BMI** is associated with a **3.5% decrease** in plasma β-carotene, with a 95% CI ranging from **−4.8% to −2.2%**.
-   A **1-unit decrease in BMI** leads to an **estimated 3.7% increase** in β-carotene concentration.
-   A **10-unit decrease in BMI** is associated with a **43% increase**, with a 95% CI ranging from **25% to 64%**.

This nonlinear interpretation stems from the exponential structure of the log-linear model: the relationship between BMI and β-carotene becomes **multiplicative**, not additive.

### Hypothesis Test for Linear Relationship Between BMI and log(β-carotene)

We test whether there is a statistically significant linear relationship between BMI and plasma β-carotene concentration on the log scale, based on the following hypotheses

-   **Null hypothesis:** $H_0:\beta_1=0$ (no relationship between BMI and log((β-carotene))

-   **Alternative hypothesis**: $H_1:\beta_1\neq0$

We use a **t-test** on the slope coefficient in the linear model. The test statistic follows a **t-distribution** with ( n - 2 ) degrees of freedom.

```{r}
# Get summary of the model
model_summary <- summary(model_log)

# Extract test statistic, degrees of freedom and p-value for the slope
t_stat <- model_summary$coefficients["bmi", "t value"]
df <- model_summary$df[2]  # residual df
p_value <- model_summary$coefficients["bmi", "Pr(>|t|)"]
cat("Test statistic (t):   ", round(t_stat, 2), "\n")
cat("Degrees of freedom:   ", df, "\n")
cat("Two-sided P-value:    ", format.pval(p_value, digits = 4), "\n")

```

Since the **P-value is far less than 0.05**, we **reject the null hypothesis** at the 5% significance level.\
This indicates that **BMI is a statistically significant predictor** of log-transformed plasma β-carotene levels.

The **negative value of the t-statistic** (t = –5.23) along with the **negative slope coefficient** (from earlier results) suggests an **inverse relationship**: as BMI increases, the expected log(β-carotene) concentration tends to decrease.

With **313 degrees of freedom**, the model has a strong basis for inference, and the **very low p-value (3.111e-07)** strengthens the evidence against the null hypothesis.

Thus, we conclude that there is a **statistically and practically significant linear relationship** between BMI and log(β-carotene).

# **2.1 Plasma β-carotene and Smoking Habits**

To investigate how smoking status relates to plasma β-carotene levels, we begin by converting the categorical variable smokstat into a factor variable with meaningful labels:

	•	**1** = Never smoker

	•	**2** = Former smoker

	•	**3** = Current smoker

We then present a frequency table showing the number of individuals in each smoking group, and calculate the mean and standard deviation of both **plasma β-carotene (ng/ml)** and **log-transformed plasma β-carotene** for each group.

These summaries help us understand both the distribution of the smoking status variable and the differences in β-carotene levels across categories.

```{r}
# Convert smokstat to factor with meaningful labels
data$smokstat <- factor(data$smokstat,
                        levels = c(1, 2, 3),
                        labels = c("Never smoker", "Former smoker", "Current smoker"))

# Frequency table
freq_table <- table(data$smokstat)
print("Frequency table for smokstat:")
print(freq_table)

# Calculate mean and SD for betaplasma and log(betaplasma) by smoking group
library(dplyr)

summary_table <- data %>%
  group_by(smokstat) %>%
  summarise(
    Count = n(),
    Mean_beta = mean(betaplasma, na.rm = TRUE),
    SD_beta = sd(betaplasma, na.rm = TRUE),
    Mean_log = mean(log(betaplasma), na.rm = TRUE),
    SD_log = sd(log(betaplasma), na.rm = TRUE)
  )

# Show the summary table
print("Summary statistics by smoking category:")
print(summary_table)
```

These summaries will later guide our modeling choices, including the selection of a reference category for regression and whether a log transformation of the response variable remains appropriate when including smokstat as a predictor.

**Reference Category Choice and Distribution Plots**

Given the three smoking categories — *Never smoker*, *Former smoker*, and *Current smoker* — we choose **Never smoker** as the reference category in our regression modeling. This choice is motivated by the following:

	•	It is the **largest group** (n = 157), ensuring stable estimates when used as baseline.

	•	It represents a **natural baseline** with respect to exposure — those who have not been exposed to smoking-related influences.

	•	Comparing other groups (former and current smokers) to this “cleanest” category allows for **straightforward interpretation** of differences in plasma β-carotene levels.

To further assess differences between groups and evaluate the need for transformation, we present boxplots of plasma β-carotene and log-transformed β-carotene across smoking status:

```{r}

# Raw scale
p_beta <- ggplot(data, aes(x = smokstat, y = betaplasma, fill = smokstat)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Plasma β-carotene by Smoking Status",
       x = "Smoking Status", y = "β-carotene (ng/ml)") +
  theme_minimal() +
  theme(legend.position = "none")

# Log scale
p_log <- ggplot(data, aes(x = smokstat, y = log(betaplasma), fill = smokstat)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Log(Plasma β-carotene) by Smoking Status",
       x = "Smoking Status", y = "log(β-carotene)") +
  theme_minimal() +
  theme(legend.position = "none")

p_beta / p_log
```

The boxplots reveal substantial skewness and variability in the raw β-carotene values, especially among non-smokers and former smokers. The **log-transformed plot** shows a more symmetric and homoscedastic distribution across groups, supporting the continued use of **log(β-carotene)** as the dependent variable in regression modeling.

## **2.2 Modeling β-carotene and Smoking Status**

**Comparing Reference Categories in Categorical Regression**

We investigate how plasma β-carotene levels (on the log scale) differ across smoking status categories by fitting two versions of a linear regression model:

	•	In **Model A**, the reference category is **“Never smoker”**

	•	In **Model B**, the reference category is **“Current smoker”**

```{r}
# Fit model with "Never smoker" as reference category
data$smokstat_never <- relevel(data$smokstat, ref = "Never smoker")
model_never <- lm(log(betaplasma) ~ smokstat_never, data = data)
summary_never <- summary(model_never)

# Fit model with "Current smoker" as reference category
data$smokstat_current <- relevel(data$smokstat, ref = "Current smoker")
model_current <- lm(log(betaplasma) ~ smokstat_current, data = data)
summary_current <- summary(model_current)

# Extract summary tables
cat("Model with 'Never smoker' as reference:\n")
summary_never$coefficients

cat("\nModel with 'Current smoker' as reference:\n")
summary_current$coefficients
```

### **Interpretation of Dummy Coding**

	•	In both models, the **intercept** represents the mean log(β-carotene) level for the **reference category**.

	•	The other coefficients represent the difference in mean log(β-carotene) compared to the reference.

	•	In Model A, the intercept corresponds to “Never smoker”, and the coefficients show how much lower (on average) former or current smokers are.

	•	In Model B, the intercept corresponds to “Current smoker”, and the coefficients show how much higher never and former smokers are.

### **Which Reference Category Is More Reasonable?**

We argue that **“Never smoker” is the more reasonable reference category**, for several reasons:

	1.	**Interpretability**: Using a baseline group that is not affected by the exposure (smoking) allows for clearer interpretation of how smoking status impacts β-carotene levels. It’s more intuitive to interpret how smoking decreases levels relative to a clean baseline (non-smoker).

	2.	**Scientific Relevance**: “Never smokers” likely represent the physiological baseline, whereas “Current smokers” are more likely to be a group with altered biology.

	3.	**Model Stability**: Model A exhibits **smaller standard errors** for the intercept and slightly more balanced standard errors overall. This suggests more stable estimation around the baseline.

As a result, we will refer to **Model A (with “Never smoker” as the reference)** as **Model 2(b)** in subsequent analyses.

## 2.3 Predicted β-carotene Levels by Smoking Group

To further understand the relationship between smoking habits and plasma β-carotene concentration, we now compute predicted values for each smoking group using both model versions:

-   **Model A**: Uses "Never smoker" as the reference category.
-   **Model B**: Uses "Current smoker" as the reference category.

For each model, we calculate:

-   The **expected log( β-carotene )** level with 95% confidence intervals.
-   The **expected β-carotene (ng/ml)** level on the original scale by back-transforming the predictions (i.e., applying the exponential function).

We compute these predictions for: - Never smokers - Former smokers - Current smokers

```{r}
# Generate a prediction data frame with each smoking category
newdata <- data.frame(smokstat = factor(c("Never smoker", "Former smoker", "Current smoker"), levels = levels(data$smokstat)))

# Model A: Never smoker as reference
model_never <- lm(log(betaplasma) ~ smokstat, data = data)

# Predict log(betaplasma) and 95% CI using Model A
pred_log_never <- predict(model_never, newdata = newdata, interval = "confidence")

# Back-transform to betaplasma scale (ng/ml)
pred_beta_never <- exp(pred_log_never)

# Model B: Current smoker as reference
data$smokstat_current <- relevel(data$smokstat, ref = "Current smoker")
model_current <- lm(log(betaplasma) ~ smokstat_current, data = data)

# Predict log(betaplasma) using Model B
newdata_current <- data.frame(smokstat_current = factor(c("Never smoker", "Former smoker", "Current smoker"), levels = levels(data$smokstat_current)))

pred_log_current <- predict(model_current, newdata = newdata_current, interval = "confidence")
pred_beta_current <- exp(pred_log_current)

# Combine results for comparison
results <- data.frame(
  Group = c("Never smoker", "Former smoker", "Current smoker"),
  
  # Log scale predictions
  Log_ModelA_Estimate = round(pred_log_never[, "fit"], 3),
  Log_ModelA_Lower    = round(pred_log_never[, "lwr"], 3),
  Log_ModelA_Upper    = round(pred_log_never[, "upr"], 3),
  
  Log_ModelB_Estimate = round(pred_log_current[, "fit"], 3),
  Log_ModelB_Lower    = round(pred_log_current[, "lwr"], 3),
  Log_ModelB_Upper    = round(pred_log_current[, "upr"], 3),
  
  # Back-transformed scale
  Beta_ModelA_Estimate = round(pred_beta_never[, "fit"], 1),
  Beta_ModelA_Lower    = round(pred_beta_never[, "lwr"], 1),
  Beta_ModelA_Upper    = round(pred_beta_never[, "upr"], 1),
  
  Beta_ModelB_Estimate = round(pred_beta_current[, "fit"], 1),
  Beta_ModelB_Lower    = round(pred_beta_current[, "lwr"], 1),
  Beta_ModelB_Upper    = round(pred_beta_current[, "upr"], 1)
)

print(results)
```

### Interpretation

We see that the predicted values and their 95% confidence intervals are **identical** for both models:

-   For **Never smokers**, the estimated plasma β-carotene level is approximately **156.2 ng/ml** with a 95% CI of **138.8 to 175.7**.
-   For **Former smokers**, the level is approximately **139.9 ng/ml** with a CI of **121.9 to 160.6**.
-   For **Current smokers**, it is around **100.9 ng/ml**, with a CI of **80.5 to 126.3**.

This confirms that **the predictions and their confidence intervals are invariant to the choice of reference level**, as expected. Changing the reference category affects the interpretation of the regression coefficients, but not the actual fitted values or predictions.

These values also align well with the group-wise means from section 2(a), reinforcing that the model provides an appropriate summary of the data.

## 2.4 Testing for Differences Between Smoking Groups

To evaluate whether smoking status has a statistically significant effect on plasma β-carotene levels (on the log scale), we fit the model:

```{r}
# Fit the linear model using log-transformed outcome
model_smok <- lm(log(betaplasma) ~ smokstat, data = data)

# Perform ANOVA (F-test)
anova_result <- anova(model_smok)

# View full ANOVA table
print(anova_result)

# Extract values for reporting
f_stat <- anova_result["smokstat", "F value"]
df1 <- anova_result["smokstat", "Df"]
df2 <- anova_result["Residuals", "Df"]
p_val <- anova_result["smokstat", "Pr(>F)"]

cat("F-statistic:", round(f_stat, 2), "\n")
cat("Degrees of freedom:", df1, "and", df2, "\n")
cat("P-value:", format.pval(p_val, digits = 4), "\n")
```

We then applied an **ANOVA F-test** to compare the group means across the three smoking categories (Never smoker, Former smoker, and Current smoker).

The hypotheses for the test are:

	•	**Null hypothesis (H₀):** μ₁ = μ₂ = μ₃ — all groups have the same mean log plasma β-carotene level.

	•	**Alternative hypothesis (H₁):** At least one group has a different mean. The test result gave the following:

	•	**F-statistic:** 5.75

	•	**Degrees of freedom:** 2 and 312

	•	**P-value:** 0.00353

### **Conclusion**

Since the p-value is below the significance level of 0.05, we **reject the null hypothesis**. This indicates that there is a statistically significant difference in mean log plasma β-carotene levels among the different smoking status categories.

# 3.1 Multiple Linear Regression
